---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  error = FALSE
)
```

# SpeciesPoolR

<!-- badges: start -->
<!-- badges: end -->

The goal of the `SpeciesPoolR` package is to generate potential species pools and their summary metrics in a spatial way.

## Installation

You can install the development version of SpeciesPoolR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("derek-corcoran-barrios/SpeciesPoolR")
```


No you can load the package

```{r setup}
library(SpeciesPoolR)
```

## Summary metrics

Now that you have in which habitat and where a species can be found, we can generate several summary metrics, we will start by reading in the resulting potential occurrence by species in.

## Read in Presences from a folder

The first thing to do is to read in Presences from a folder, with the function `GetLandusePresences` as seen in the following code

```{r GetLandusePresences}

folder <- system.file("ex/", package="SpeciesPoolR")
Landuse <- "ForestDryPoor"
presences <- GetLandusePresences(folder, Landuse)
```

This is the first 10 lines of the file

```{r Table1, echo=FALSE}
knitr::kable(head(presences))
```
