---
title: "How to use the SpeciesPoolR package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use the SpeciesPoolR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of the `SpeciesPoolR` package is to generate potential species pools and their summary metrics in a spatial way, to install the package you can do it from github:

```{r, eval=FALSE}
#install.packages("remotes")
remotes::install_github("derek-corcoran-barrios/SpeciesPoolR")
```

No you can load the package

```{r setup}
library(SpeciesPoolR)
```

# Summary metrics

Now that you have in which habitat and where a species can be found, we can generate several summary metrics, we will start by reading in the resulting potential occurrence by species in.

## Read in Presences from a folder

The first thing to do is to read in Presences from a folder, with the function `GetLandusePresences` as seen in the following code

```{r GetLandusePresences}

folder <- system.file("ex/", package="SpeciesPoolR")
Landuse <- "ForestDryPoor"
presences <- GetLandusePresences(folder, Landuse)
```

This is the first 10 lines of the file

```{r Table1, echo=FALSE}
knitr::kable(head(presences))
```

