---
title: "How to use the SpeciesPoolR package"
output:
  bookdown::html_vignette2:
    fig_caption: true
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{How to use the SpeciesPoolR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: 'Biblio.bib'
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# SpeciesPoolR

<!-- badges: start -->
[![R-CMD-check](https://github.com/derek-corcoran-barrios/SpeciesPoolR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/derek-corcoran-barrios/SpeciesPoolR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of the SpeciesPoolR package is to generate potential species pools and their summary metrics in a spatial way. You can install the package directly from GitHub:

```{r, eval=FALSE}
#install.packages("remotes")
remotes::install_github("derek-corcoran-barrios/SpeciesPoolR")
```

No you can load the package

```{r setup}
library(SpeciesPoolR)
```

# Using SpeciesPoolR Manually

## Importing and Downloading Species Presences

### Step 1: Reading and Filtering Data

If you are going to use each of the functions of the SpeciesPoolR manually and sequentially, the first step would be to read in a species list from either a CSV or an XLSX file. You can use the get_data function for this. The function allows you to filter your data in a dplyr-like style:

```{r ReadInData}
f <- system.file("ex/Species_List.csv", package="SpeciesPoolR")
filtered_data <- get_data(
   file = f,
   filter = quote(Kingdom == "Plantae" & 
                    Class == "Magnoliopsida" & 
                    Family == "Fabaceae")
)
```

This will generate a dataset that can be used subsequently to count species presences and download species data as seen in table \@ref(tab:tablespecies)

```{r tablespecies, echo = F}
knitr::kable(filtered_data, caption = "Species that will be used to generate species pools")
```

### Step 2: Taxonomic Harmonization

Next, you should perform taxonomic harmonization to ensure that the species names you use are recognized by the GBIF taxonomic backbone. This can be done using the Clean_Taxa function:

```{r CleanSpecies}
Clean_Species <- SpeciesPoolR::Clean_Taxa(filtered_data$Species)
```

The resulting data frame, with harmonized species names, is shown in table \@ref(tab:cleantable)

```{r cleantable, echo=FALSE}
knitr::kable(Clean_Species, caption = "Taxonomicallty harmonized dataset")
```

### Step 3: Counting Species Presences

After harmonizing the species names, it's important to obtain the number of occurrences of each species in your study area, especially if you plan to calculate rarity. You can do this using the `count_presences` function. This function allows you to filter occurrences by country or by a shapefile. Below is an example for Denmark:

```{r countdenmark}
# Assuming Clean_Species is your data frame
Count_DK <- count_presences(Clean_Species, country = "DK")
```

The resulting data frame of species presences in Denmark is shown in table \@ref(tab:tableCountDenmark)

```{r tableCountDenmark}
knitr::kable(Count_DK, caption = "Counts of presences for the different species within Denmark")
```

Alternatively, you can filter by a specific region using a shapefile. For example, to count species presences within Aarhus commune:

```{r AarhusCount}
shp <- system.file("ex/Aarhus.shp", package="SpeciesPoolR")

Count_Aarhus <- count_presences(Clean_Species, shapefile = shp)
```

The resulting data.frame for Aarhus commune is shown int table \@ref(tab:tableCountAarhus)

```{r tableCountAarhus, echo = FALSE}
knitr::kable(Count_Aarhus, caption = "Counts of presences for the different species within Aarhus commune")
```

Now it is recommended to eliminate species that have no occurrences in the area, this is done automatically in the workflow version:

```{r filter}
library(data.table)
Count_Aarhus <- Count_Aarhus[N > 0,]
```

So that then we can retrieve the species presences using the function `get_presences`.

```{r getPresences}
Presences <- get_presences(species = Count_Aarhus$species, shapefile = shp)
```

there we end up with `r nrow(Presences)` presences for our `r nrow(Count_Aarhus)` species.

## Creating Spatial Buffers and Habitat Filtering

### Step 1 Creating Buffers Around Species Presences

Once you have obtained the species presences within your area of interest, the next step is to create spatial buffers around these occurrences. These buffers simulate the potential dispersal range of each species. By doing so, you can assess the possible areas where each species could establish itself, considering a given dispersal distance.

To create these buffers, you'll need a raster file that serves as a template for rasterizing the buffers. You'll also specify the distance (in meters) that represents the dispersal range of the species.

```{r buffer500}
Raster <- system.file("ex/LU_Aarhus.tif", package="SpeciesPoolR")

buffer500 <- make_buffer_rasterized(Presences, file = Raster, dist = 500)
```

In the example above, the `make_buffer_rasterized` function creates a 500-meter buffer around each occurrence point of the species in the Presences dataset. The function uses the provided raster file as a template to rasterize the buffers.

The resulting data frame, `buffer500`, contains information on which raster cells are covered by the buffer for each species. The first 10 observations of this data frame are shown in Table \@ref(tab:showbuffer500).

```{r showbuffer500, echo=FALSE}
knitr::kable(head(buffer500, 10), caption = "Raster cells within the 500-meter buffer of each species")
```

This table provides a detailed view of how the buffer overlaps with the raster cells, listing each cell and the corresponding species present within that buffer.

### Step 2: Habitat Filtering

After creating the buffers, you may want to filter the areas further based on habitat suitability. For instance, you might only be interested in specific land-use types or habitats where the species is likely to thrive. This step typically involves using the raster data to mask or subset the buffer areas according to the desired habitat criteria.

For that we will use the `ModelAndPredictFunc` which will take the data frame of presences obtained through the `get_presences` and modeling through the landuse raster that we have as an example in the dataset:

```{r habitatfiltering}
Habitats <- ModelAndPredictFunc(DF = Presences, file = Raster)
```


## Generating summary biodiversity statistics

### Step 1 Generating Phylogenetic diversity metrics

In order to generate Phylogenetic Diversity measures, the first step is to generate a phylogenetic tree with the species we have, for that we will use the V.Phylomaker package function `phylo.maker`based on the megaphylogeny of vascular plants [@Jin2019; @Zanne2014], this means that we can only use this functions in species pools of plants.

In this case we use the `generate_tree` from SpeciesPoolR to do so:

```{r treegeneration}
tree <- generate_tree(Count_Aarhus)
```

# Running the SpeciesPoolR Workflow

If you prefer to automate the process and run the `SpeciesPoolR` workflow as a pipeline, you can use the `run_workflow` function. This function sets up a `targets` workflow that sequentially executes the steps for cleaning species data, counting species presences, and performing spatial analysis. This approach is especially useful for larger datasets or when you want to ensure reproducibility.

To run the workflow, you can use the following code. We'll use the same species filter as before, focusing on the `Plantae` kingdom, `Magnoliopsida` class, and `Fabaceae` family. Additionally, we'll focus on the Aarhus commune using a shapefile.

```{r run_workflow}
shp <- system.file("ex/Aarhus.shp", package = "SpeciesPoolR")
Raster <- system.file("ex/LU_Aarhus.tif", package="SpeciesPoolR")

run_workflow(
  file_path = system.file("ex/Species_List.csv", package = "SpeciesPoolR"),
  filter = quote(Kingdom == "Plantae" & Class == "Magnoliopsida" & Family == "Fabaceae"),
  shapefile = shp,
  dist = 500,
  rastertemp = Raster
)
```


## How It Works

The run_workflow function creates a pipeline that:

1- Reads the data from the specified file path.
2- Filters the data using the provided filter expression.
3- Cleans the species names to match the GBIF taxonomic backbone.
4- Counts the species presences within the specified geographic area (in this case, Aarhus).
5- Generates a buffer around the species presences within the specified distance, for a template raster.
6- Generates a phyllogenetic tree for the species in the species list.
7- Generates a visual representation of the workflow (if plot = TRUE).

You can monitor the progress of the workflow and visualize the dependencies between steps using targets::tar_visnetwork(). The result will be similar to running the steps manually but with the added benefits of parallel execution and reproducibility.

This automated approach allows you to streamline your analysis and ensures that all steps are consistently applied to your data. It also makes it easier to rerun the workflow with different parameters or datasets.

```{r removefiles, include=FALSE}
unlink("_targets/", recursive = T, force = T)
file.remove("_targets.R")
```
